---
title: "Final Project"
author: "Max Dodge"
date: "5/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries

```{r, message=FALSE, warning=FALSE}
library(readr)
library(ggplot2)
library(dplyr)
library(caret)
library(lubridate)
library(tidyr)
library(zoo)
library(ISLR)
library(splines)
library(gam)
library(rpart)
library(rpart.plot)
```

### Data

```{r}
us_counties <- #Kaggle dataset originally from the NYT
  read_csv("us-counties.csv")

census_pop <- 
  read_csv("CensusCountyPop.csv")

mask_use <- 
  read_csv("mask-use-by-county.csv")  #Kaggle dataset

pfizer <- 
  read_csv("COVID-19_Vaccine_Distribution_Allocations_by_Jurisdiction_-_Pfizer (1).csv", 
    col_types = cols(`Week of Allocations` = col_date(format = "%m/%d/%Y"))) #CDC data

moderna <- 
  read_csv("COVID-19_Vaccine_Distribution_Allocations_by_Jurisdiction_-_Moderna.csv", 
    col_types = cols(`Week of Allocations` = col_date(format = "%m/%d/%Y"))) #CDC data

janssen <- 
  read_csv("COVID-19_Vaccine_Distribution_Allocations_by_Jurisdiction_-_Janssen.csv", 
    col_types = cols(`Week of Allocations` = col_date(format = "%m/%d/%Y"))) #CDC data
```

### Data Cleaning

```{r}
state_pop <- census_pop %>% 
  filter(COUNTY == "000") %>% 
  select(STNAME, POPESTIMATE2019)

county_pop <- census_pop %>%
  select(STATE,COUNTY,STNAME,CTYNAME,POPESTIMATE2019) %>%
  mutate(fips = paste0(STATE, COUNTY)) %>% 
  select(POPESTIMATE2019, fips)
```

```{r}
covid_data <- us_counties %>% 
  left_join(mask_use, by = c("fips" = "COUNTYFP"))

covid_data <- covid_data %>% 
  left_join(county_pop, by = "fips") %>% 
  na.omit()

covid_data <- covid_data %>% 
  left_join(state_pop, by = c("state" = "STNAME")) %>% 
  na.omit() %>% 
  rename(pop_county = POPESTIMATE2019.x,
         pop_state = POPESTIMATE2019.y)

covid_data <- covid_data %>% 
  left_join(janssen, by = (c("date" = "Week of Allocations", 
                             "state" = "Jurisdiction"))) %>% 
  rename(janssen_doses = `1st Dose Allocations`) %>% 
  replace_na(list(janssen_doses = 0)) %>% 
  left_join(pfizer, by = (c("date" = "Week of Allocations", 
                             "state" = "Jurisdiction"))) %>% 
  rename(pfizer_dose_1 = `1st Dose Allocations`,
         pfizer_dose_2 = `2nd Dose Allocations`) %>% 
  replace_na(list(pfizer_dose_1 = 0, pfizer_dose_2 = 0)) %>% 
  left_join(moderna, by = (c("date" = "Week of Allocations", 
                             "state" = "Jurisdiction"))) %>% 
  rename(moderna_dose_1 = `1st Dose Allocations`,
         moderna_dose_2 = `2nd Dose Allocations`) %>% 
  replace_na(list(moderna_dose_1 = 0, moderna_dose_2 = 0))

```

### Creating New Variables

```{r}
covid_data <- covid_data %>% 
  mutate(total_doses = janssen_doses + pfizer_dose_1 + pfizer_dose_2 + 
           moderna_dose_1 + moderna_dose_2)

covid_data <- covid_data %>%
  filter(date > "2020-11-30") %>% 
  mutate(prop_statepop = pop_county/pop_state) %>% 
  mutate(estimated_doses = prop_statepop*total_doses) %>% 
  mutate(cases_per10k = (10000*cases)/(pop_county))

covid_data <- covid_data %>% 
  mutate(severe_outbreak = (ifelse(cases_per10k >1000,"Severe","Not_Severe"))) 
  
```

## Regression

### Ordinary Least Squares Model

```{r, warning=FALSE, message=FALSE}
set.seed(2021)

OLS <- train(
    cases_per10k ~ date+NEVER+ALWAYS+SOMETIMES+RARELY+FREQUENTLY+janssen_doses+pfizer_dose_1+pfizer_dose_2+moderna_dose_1+moderna_dose_2+total_doses+pop_county+estimated_doses, 
    data = covid_data,
    method = "lm",
    trControl = trainControl(method = "cv", number = 10),
    na.action = na.omit)

```

### Ordinary Least Squares Analysis

```{r, warning=FALSE, message=FALSE}
OLS_out <- covid_data %>% 
  mutate(fitted = predict(OLS, newdata = covid_data),
         resid = cases_per10k - fitted)

```

```{r}
OLS_out %>% 
  ggplot(aes(x = fitted, y = resid)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Fitted values", y = "Residuals")

```

```{r}
OLS_out %>% 
  ggplot(aes(x = total_doses, y = resid)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Alotted Total Doses", y = "Residuals")

```

```{r}
OLS_out %>% 
  ggplot(aes(x = date, y = resid)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Date", y = "Residuals")

```

```{r}
head(OLS$results %>% arrange(MAE),1)
```

```{r}
coef(OLS$finalModel)
```

### Linear LASSO Model

```{r}
set.seed(2021)

LASSO <- train(
    cases_per10k ~ date+NEVER+ALWAYS+SOMETIMES+RARELY+FREQUENTLY+janssen_doses+pfizer_dose_1+
      pfizer_dose_2+moderna_dose_1+moderna_dose_2+total_doses+pop_county+estimated_doses,
    data = covid_data,
    method = "glmnet",
    tuneGrid = data.frame(alpha = 1, lambda = seq(0, 10, length.out = 100)),
    trControl = trainControl(method = "cv", number = 10, selectionFunction = "oneSE"),
    metric = "MAE",
    na.action = na.omit)

```

### Linear LASSO Analysis

```{r}
LASSO_out <- covid_data %>% 
  mutate(fitted = predict(LASSO, newdata = covid_data),
         resid = cases_per10k - fitted)

```

```{r}
LASSO_out %>% 
  ggplot(aes(x = fitted, y = resid)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Fitted values", y = "Residuals")

```

```{r}
LASSO_out %>% 
  ggplot(aes(x = total_doses, y = resid)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Alotted Total Doses", y = "Residuals")

```

```{r}
coef(LASSO$finalModel, LASSO$bestTune$lambda)
```

```{r}
head(LASSO$results %>% arrange(MAE),1)
```

```{r}
plot(LASSO)
```

### Linear Forward Stepwise Model

```{r, warning=FALSE, message=FALSE}
set.seed(2021)

ForwardSW <- train(
    cases_per10k ~ date+NEVER+ALWAYS+SOMETIMES+RARELY+FREQUENTLY+janssen_doses+pfizer_dose_1+
      pfizer_dose_2+moderna_dose_1+moderna_dose_2+total_doses+pop_county+estimated_doses,
    data = covid_data,
    method = "leapForward",
    tuneGrid = data.frame(nvmax = 1:10),
    trControl = trainControl(method = "cv", number = 10, selectionFunction = "oneSE"),
    metric = "MAE",
    na.action = na.omit)
```

### Linear Forward Stepwise Analysis

```{r}
ForwardSW_out <- covid_data %>% 
  mutate(fitted = predict(ForwardSW, newdata = covid_data),
         resid = cases_per10k - fitted)

```

```{r}
ForwardSW_out %>% 
  ggplot(aes(x = fitted, y = resid)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Fitted values", y = "Residuals")

```

```{r}
ForwardSW_out %>% 
  ggplot(aes(x = total_doses, y = resid)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Alotted Total Doses", y = "Residuals")

```

```{r}
head(ForwardSW$results %>% arrange(MAE),1)
```

```{r}
coef(ForwardSW$finalModel, ForwardSW$bestTune$nvmax)
```

### LASSO with Splines Model

```{r}
cs_lasso <- train(
    cases_per10k ~ ns(date, 3)+ns(NEVER, 3)+ns(RARELY, 3)+ns(ALWAYS, 3)+ns(FREQUENTLY, 3)+
      ns(pfizer_dose_1, 3)+ns(pfizer_dose_2, 3)+ns(janssen_doses, 3)+ns(SOMETIMES, 3)+
      ns(moderna_dose_1, 3)+ns(moderna_dose_2, 3)+ns(pop_county, 3)+ns(total_doses, 3)+ns(estimated_doses, 3),
    data = covid_data,
    method = "glmnet",
    tuneGrid = data.frame(alpha = 1, lambda = seq(0, 10, length.out = 100)),
    trControl = trainControl(method = "cv", number = 10, selectionFunction = "oneSE"),
    metric = "MAE",
    na.action = na.omit)

```

### LASSO with Splines Analysis

```{r}
cs_lasso_out <- covid_data %>% 
  mutate(fitted = predict(cs_lasso, newdata = covid_data),
         resid = cases_per10k - fitted)

```

```{r}
cs_lasso_out %>% 
  ggplot(aes(x = fitted, y = resid)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Fitted values", y = "Residuals")
```

```{r}
cs_lasso_out %>% 
  ggplot(aes(x = total_doses, y = resid)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Alotted Total Doses", y = "Residuals")
```

```{r}
head(cs_lasso$results %>% arrange(MAE),1)
```

```{r}
plot(cs_lasso)
```

```{r}
coef(cs_lasso$finalModel, LASSO$bestTune$lambda)
```


### Forward Stepwise with Splines Model

```{r, warning=FALSE, message=FALSE}
set.seed(2021)

cs_forward <- train(
    cases_per10k ~  ns(date, 3)+ns(NEVER, 3)+ns(RARELY, 3)+ns(ALWAYS, 3)+ns(FREQUENTLY, 3)+
      ns(pfizer_dose_1, 3)+ns(pfizer_dose_2, 3)+ns(janssen_doses, 3)+ns(SOMETIMES, 3)+
      ns(moderna_dose_1, 3)+ns(moderna_dose_2, 3)+ns(pop_county, 3)+ns(total_doses, 3)+ns(estimated_doses, 3),
    data = covid_data,
    method = "leapForward",
    tuneGrid = data.frame(nvmax = 1:10),
    trControl = trainControl(method = "cv", number = 10, selectionFunction = "oneSE"),
    metric = "MAE",
    na.action = na.omit)

```

### Forward Stepwise with Splines Analysis

```{r}
cs_forward_out <- covid_data %>% 
  mutate(fitted = predict(cs_forward, newdata = covid_data),
         resid = cases_per10k - fitted)

```

```{r}
cs_forward_out %>% 
  ggplot(aes(x = fitted, y = resid)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Fitted values", y = "Residuals")

```

```{r}
cs_forward_out %>% 
  ggplot(aes(x = total_doses, y = resid)) +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(x = "Alotted Total Doses", y = "Residuals")

```

```{r}
head(cs_forward$results %>% arrange(MAE),1)
```

```{r}
coef(cs_forward$finalModel, ForwardSW$bestTune$nvmax)
```

## Classification

### LASSO Logistic Regression Model

```{r}
twoClassSummaryCustom <- function (data, lev = NULL, model = NULL) {
    if (length(lev) > 2) {
        stop(paste("Your outcome has", length(lev), "levels. The twoClassSummary() function isn't appropriate."))
    }
    caret:::requireNamespaceQuietStop("pROC")
    if (!all(levels(data[, "pred"]) == lev)) {
        stop("levels of observed and predicted data do not match")
    }
    rocObject <- try(pROC::roc(data$obs, data[, lev[1]], direction = ">", 
        quiet = TRUE), silent = TRUE)
    rocAUC <- if (inherits(rocObject, "try-error")) 
        NA
    else rocObject$auc
    out <- c(rocAUC, sensitivity(data[, "pred"], data[, "obs"], 
        lev[1]), specificity(data[, "pred"], data[, "obs"], lev[2]))
    out2 <- postResample(data[, "pred"], data[, "obs"])
    out <- c(out, out2[1])
    names(out) <- c("AUC", "Sens", "Spec", "Accuracy")
    out
} # This function is not my own, from exercise 10 (Evaluating Classification Models)

set.seed(2021)

lasso_log <- train(
    severe_outbreak ~ date+NEVER+ALWAYS+SOMETIMES+
      RARELY+FREQUENTLY+janssen_doses+pfizer_dose_1+
      pfizer_dose_2+moderna_dose_1+moderna_dose_2+
      total_doses+pop_county+estimated_doses,   
    data = covid_data,
    method = "glmnet",
    family = "binomial",
    tuneGrid = data.frame(alpha = 1, lambda = seq(0, 10, length.out = 100)),
    trControl = trainControl(method = "cv", number = 10, selectionFunction = "oneSE", classProbs = TRUE, summaryFunction = twoClassSummaryCustom),
    metric = "AUC",
    na.action = na.omit
)

```

### LASSO Logistic Regression Analyisis

```{r}
coef(lasso_log$finalModel, lasso_log$bestTune$lambda)
```

```{r}
lasso_log$results %>%
    filter(lambda==lasso_log$bestTune$lambda)
```

```{r}
covid_data %>%
    count(severe_outbreak)
```

```{r}
# NIR
298643/(92973 + 298643)
```

### Random Forest Model

```{r}
set.seed(2021)
# This code takes a really long time(for me ~7 minutes)
forest <- train(
    severe_outbreak ~ date+NEVER+ALWAYS+SOMETIMES+
      RARELY+FREQUENTLY+janssen_doses+pfizer_dose_1+
      pfizer_dose_2+moderna_dose_1+moderna_dose_2+
      total_doses+pop_county+estimated_doses,
    data = covid_data,
    method = "rf",
    tuneGrid = data.frame(mtry = c(2,3,4,5)),
    trControl = trainControl(method = "oob", selectionFunction = "best"),
    metric = "Accuracy",
    ntree = 50, 
    na.action = na.omit
)
```

### Random Forest Analysis

```{r}
plot(forest)
```

```{r}
forest$results
```

```{r}
forest$finalModel
```

```{r}
variable_importance <- randomForest::importance(forest$finalModel)

variable_importance <- data.frame(
        predictor = rownames(variable_importance),
        MeanDecreaseGini = variable_importance[,"MeanDecreaseGini"]
    ) %>%
    arrange(desc(MeanDecreaseGini))

variable_importance
```

